---
title: "Design and Analysis of Experiments"
author: "David"
date: '`r Sys.Date()`'
output: html_document
---

```{r}
# load packages
suppressMessages(
  {
    library(tidyverse)
    library(agricolae)
    library(report)
  }
)
```


# 1. Analysis of a Completely Randomised Design (CRD)

```{r}
## Comparing if different foodstuffs given to chicks significantly affect their weights

# Create a vector containing the weights of chicks
weights <- c(55,49,42,21,52,61,112,30,89,63,
             42,97,81,95,92,169,137,169,85,154)

# create another vector containing the various foodstuffs. 
Foodstuffs <- c("A","A","A","A","A","B","B","B","B","B",
                "C","C","C","C","C","D","D","D","D","D")

# create a data frame of the weights and foodstuffs
Foodstuff_data <- data.frame(weights, Foodstuffs)

# fit a CRD model
CRD <- aov(weights ~ Foodstuffs, data = Foodstuff_data)

# model summary
summary(CRD)
```

Since the p-value(0.000218) is much less than alpha(0.05), I reject Ho at 5% level of significance and conclude that the foodstuffs differ significantly with regards to increase in weight of chicks.

```{r}
# carry out post-hoc multiple comparison test using Least Significant Difference test
d <- LSD.test(CRD, "Foodstuffs", DFerror = 16, MSerror = 722, 
              alpha = 0.05, console = TRUE)

# print LSD test results
d
```

Foodstuffs C and B have almost the same effect, same as B and A. D is statistically different from the other foodstuffs. D has the highest mean weight, hence it is the best and is the one which should be fed to the chicks.

```{r}
# visualize the result
plot(d, las = 1, horiz = TRUE)
```

From this distribution, weights of chicks which were fed on foodstuff D varies differently from the others.

```{r}
# use Tukey's HSD test
HSD <- TukeyHSD(CRD, conf.level = .95)
HSD
```

Foodstuffs D and A are significantly different, same as D and B and also D and C, since their p-values are less than 0.05. I therefore conclude that foodstuff D is significantly different from C, B and A.

```{r}
# visualize the confidence intervals
plot(HSD, las = 1)
```

This distribution shows that B&A, C&A, C&B differ from D&A, D&B, D&C.

```{r}
## test whether the mean sales volume for 3 new menu items are all equal.

# first create the data
Sales <- c(31,31,45,21,42,32,27,28,29,18,36,17,24,31,46,48,46,40)
Items <- rep(c("Item1","Item2","Item3"), each = 6)
Item_sales <- data.frame(Sales, Items)

# view the first few observations of the Item_Sales data set
head(Item_sales)

# fit ANOVA model
Items_model <- aov(Sales ~ Items, data = Item_sales)
# model summary
summary(Items_model)
```

Since the p-value(0.0506) is greater than alpha(0.05), I fail to reject Ho at 5% level of significance and conclude that the mean sales volume for the 3 new menu items are all equal. (Unless there's a statistical power that says otherwise.)

```{r}
## Analysis of a CRD and use of Duncan's multiple range test and Student-Newman-Keuls(SNK) test for post hoc tests.

# load the sweet potato data set
data("sweetpotato")

# view the structure of the data set
glimpse(sweetpotato)
```

The sweet potato dataset contains 12 observations of 2 variables. Virus is a categorical (factor) variable while yield is numeric.

```{r}
# clean the data

# check if the data has any missing values
map_dbl(sweetpotato, ~sum(which(is.na(.))))

# check for duplicated observations
sum(duplicated(sweetpotato))
```

The data is clean, has no missing values or duplicates.

```{r}
# carry out ANOVA to assess whether the viruses had the same effect on potato yield
sp_model <- aov(yield ~ virus, data = sweetpotato)
# model summary
summary(sp_model)
```

Since the p-value = 0.000733 is much less than alpha(0.05), I reject Ho at 5% significance level and conclude that the viruses have significantly different effects on potato yield.

```{r}
# carry out post-hoc multiple comparison test using Duncan's new multiple range test
dcun_test <- duncan.test(sp_model, "virus", alpha = 0.05, 
                         group = TRUE, console = TRUE)
dcun_test
```

Viruses oo and ff have almost the same effect on potato yield but their effects significantly differ from the effects of viruses cc and fc.

```{r}
# visualize the variations in yields
plot(dcun_test, horiz = TRUE, las = 1)
plot(dcun_test, variation = "IQR", horiz = TRUE)
```


The variations and IQR pots also show that viruses oo and ff have almost the same effect on sweet potato yields, and differs significantly from viruses cc and fc.

```{r}
# use Student-Newman-Keuls(SNK) test
snk_test <-  SNK.test(sp_model, "virus", console = TRUE, 
                      main =  "Yield of sweetpotato dealt with different virus")
snk_test

# visualize SNK.test results
plot(snk_test, las = 1, horiz = TRUE)
```

SNK test gives the same results as Duncan's multiple range test.


# 2. Analysis of Randomised Block Design (RBD)

```{r}
## Given data on the quantity of fertilizer used during planting and the amount of water sprayed for the crops. Test if there's significant difference in yield produced

# create a vector containing fertilizer quantities
Fertilizer <- c("low","high","low","high")
# create a vector containing quantities of water applied
Irrigation <- rep(c("low","high"), each = 2)
# create a vector containing yield quantities
Yield <- c(20,40,30,52)

# create a data frame for the yields, fertilizer and irrigation
YieldData <- data.frame(Yield, Fertilizer, Irrigation)
```

```{r}
# fit the model
YieldModel <- aov(Yield ~ Fertilizer + Irrigation, data = YieldData)
# model summary
summary(YieldModel)
```

* The Fertilizer quantity applied had a significant effect on crop yield (p-value < 0.05)
* Irrigation (quantity of water applied) on the other hand had no significant effect on crop yield (p-value > 0.05)

I used the irrigation quantity as my blocks. I won't carry post-hoc analysis for this example.

```{r}
## Another example on RBD

# first create the dataset

# create a vector of values
Values <- c(24.7,20.6,27.7,16.2,16.2,24.9,27.3,28.8,22.7,15.0,17.0,22.5,
            38.5,39.5,36.8,19.6,15.4,26.3,28.5,31.0,34.9,14.1,17.7,22.6)

# create a vector of treatments
Treatments <- rep(c("T1","T2","T3","T4","T5","T6"), 4)
# create a vector of blocks
Blocks <- rep(c("B1","B2","B3","B4"), each = 6)

# combine the vectors into a data frame
RBD_data <- data.frame(Values, Treatments, Blocks)
# view the first few obs
head(RBD_data)

# fit the model
RBD <- aov(Values ~ Treatments + Blocks, data = RBD_data)
# model summary
summary(RBD)
```

* Since the p-value associated with treatments is much less than alpha(0.05), I reject Ho at 5% significance level and conclude that the treatments differ significantly.
* Since the p-value associated with blocks is also less than beta(0.05), I reject Ho at 5% significance level and conclude that the blocks differ significantly. This makes sense because in a Randomized Block Design, the blocks should be homogeneous within themselves but heterogeneous among themselves.

```{r}
# carry out LSD-test on treatments
d2 <- LSD.test(RBD, "Treatments", DFerror = 15, MSerror = 15.31, 
               alpha = 0.05, group = TRUE, console = TRUE)

# view LSD-test results
d2
```

* Paired treatments 2&3, 1&2, 1&6, 2&6, 4&5 have almost the same effect.
* Paired treatments 3&5, 3&4, 2&5, 2&4, 1&5, 1&4, 6&5 and 6&4 are significantly different.


# 3. Analysis of Latin Square Design (LSD)

```{r}

## Given productivity data based on 5 different kinds of fertilizer, 5 kinds of tillage, and 5 kinds of seeds. Test if there are significant differences between fertilizer, tillage and seed types, with regards to the quantity of yield produced. The data is organized in a Latin square design.

# First create a data frame with all the attributes
Fertilizer <- rep(c("F1","F2","F3","F4","F5"),5)

Tillage <- rep(c("A","B","C","D","E"), each = 5)

Seed <- c("a","e","c","b","d","c","b","a","d","e",
          "b","c","d","e","a","d","a","e","c","b",
          "e","d","b","a","c")

Yield <- c(42,45,41,56,47,47,54,46,52,49,55,52,57,
           49,45,51,44,47,50,54,44,50,48,43,46)

Production_data <- data.frame(Tillage, Fertilizer, Seed, Yield)
```

```{r}
# create matrices representing the attributes

# Treatments
matrix(Production_data$Tillage,5,5)
```
```{r}
# Fertilizer types
matrix(Production_data$Fertilizer,5,5)
```
```{r}
# Seed types
matrix(Production_data$Seed,5,5)
```
```{r}
# Yields
matrix(Production_data$Yield,5,5)
```

```{r}
# carry out anova
Latin_sq_design <- aov(Yield ~ Tillage + Fertilizer + Seed, data = Production_data)
# model summary
summary(Latin_sq_design)
```

* Since the p-value associated with tillage is less than 0.05, I reject Ho at 5% level of significance and conclude that the different kinds of tillage used were significantly different with regards to production yields.
* Since the p-value associated with Fertilizers is highly greater than 0.05, I fail to reject Ho at 5% level of significance and conclude that the fertilizers do not differ significantly with regards to production yields.
* Since the p-value associated with Seeds is much less than 0.05, I reject Ho at 5% significance level and conclude that the seed types differ significantly with regards to production yields.

```{r}
# carry out post-hoc multiple comparison test on Tillage and seed types using Tukey's HSD test
TukeyHSD(Latin_sq_design, conf.level=.95)
```

* Based on the Tukey's HSD test:
  * Tillage C&A, E&C differ significantly.
  * Seed types b&a, d&a, c&b, e&b also differ significantly.
  * All the fertilizer types had almost the same effect with regards to the quantity of yields produced.


# 4. Analysis of Factorial Designs (Designs with main and interaction effects)

```{r}
# Load the Tooth Growth data
data("ToothGrowth")

# View structure of the data
ToothGrowth |> glimpse()
```

The data has 60 observations of 3 variables. Length and dose are numeric while supplement is a factor with two levels.

```{r}
# Factor the variable dose
ToothGrowth[["dose"]] <- factor(ToothGrowth[["dose"]], 
                                levels = c(0.5, 1, 2), 
                                labels = c("low","medium","high") )
```

```{r}
# Quickly explore the data
ToothGrowth |> skimr::skim()
```

There are no missing values in the data. The variable supplement has 2 levels while dosage has 3 levels. The factor variables are evenly distributed. Tooth length has a mean and standard deviation of 18.81 and 7.65 respectively.

```{r}
# Visualize the distribution of tooth length, characterized by supplement and dosage
boxplot(len ~ supp * dose, data = ToothGrowth, 
        ylab="Tooth Length", 
        main="Boxplots of Tooth Growth Data")

```

Generally, there seems to be differences in the means of tooth length based on the dosage and supplement types, except for the supplement given at high dosage.

```{r}
# interaction plot
with(ToothGrowth, interaction.plot(x.factor = dose, 
                                   trace.factor = supp, response = len, 
                                   col = c("red","blue")))
```

* There seems to be no difference between supplements at high dose.
* There seems to be a main effect of dose i.e higher dose results in higher tooth length
* Also, there seems to be a main effect of supplement i.e there is a difference in tooth growth between VC and OJ supplements at low and medium dosage.

```{r}
# fit ANOVA Model
ToothGrowth_model <- aov(len ~ supp * dose, data = ToothGrowth)
summary(ToothGrowth_model)
```

The main effects of supplement and dosage are all significant, same as the interaction effect between supplement and dosage. (p < 0.05)

```{r}
# carry out post-hoc analysis using Tukey's HSD test
TukeyHSD(ToothGrowth_model)
```

* All the simple main effects are statistically significant (p-value < 0.05)
* All the interaction effects are statistically significant, except (VC:medium-OJ:low), (OJ:high-OJ:medium) , (VC:high-OJ:medium) and (VC:high-OJ:high).

```{r}
# visualize the difference in means and Confidence Intervals
plot(TukeyHSD(ToothGrowth_model))
```

```{r}
## Assessing the effects of material type and temperature on battery life

# creating the data

# create a vector of values for battery life
Values <- c(130,74,155,180,34,80,40,75,20,82,70,58,150,159,
            188,126,136,106,122,115,25,58,70,45,138,168,110,
            160,174,150,120,139,96,82,104,60)

# create a vector containing the different material types
Material_type <- rep(c("1","2","3"), each = 12)

# create a vector containing the different temperature levels
Temp <- rep(c("15","15","15","15","70","70","70","70","125","125",
              "125","125"),3)

# combine the vectors into a data frame
Battery_data <- data.frame(Values, Material_type, Temp)
```

```{r}
# fit the model
BattModel <- aov(Values ~ Material_type*Temp, data = Battery_data)
# model summary
summary(BattModel)
```

The main effects of material type and temperature are statistically significant at 5% level of significance, same as the interaction effect between material and temperature (p < 0.05).

```{r}
## Model interpretation by the report() function from report package
library(report)
BattModel |> report()
```
```{r}
# plot the interaction curve
# first convert the character variables to factors
df <- data.frame(Values, factor(Material_type), factor(Battery_data$Temp, levels = c(15,70,125)))

# plot
interaction.plot(
  x.factor = df$factor.Battery_data.Temp..levels...c.15..70..125..,
  trace.factor = df$factor.Material_type.,
  response = df$Values,
  fun = median,
  ylab = "Average battery life",
  xlab = "Temperature Level",
  trace.label = "Materia Type",
  col = c("#0198f9","#f95801","#f92676"),
  lwd = 3
)
```

* For material type 1, increase in temperature from 15 to 70 reduces the average battery life steadily from 140 to about 58. Further increase in temperature from 70 to 125 slightly increases the average battery life from 58 to about 62.
* For material type 2, increase in temperature from 15 all the way to 125 reduces the average battery life from about 155 to 50.
* For material type 3, increase in temperature from 15 to 70 has a very little effect on the average battery life. Further increase in temperature from 70 to 125 reduces the average battery life from about 150 to 90.

I can say that the best combination that guarantees a longer average battery life are; a temperature of 15 and material type 2.

```{r}
# post-hoc multiple range test using Tukey's HSD test
TukeyHSD(BattModel)
```

* There are significant differences between material types 3&1.
* There is no significant difference between material types 1&2 and 2&3.
* There are significant differences between all the 3 temperature levels.
* And just as I saw from the interaction plot, material type 2 at temperature 15 vs material type 2 at temperature 125 has the smallest p-value, hence is highly significant. This implies that using material type 2 at temperature 15 produces batteries with the best life span while using material type 2 at temperature 125 produces batteries with the shortest  life span.

```{r}
## Assessing the effects of reactant and catalyst quantities on yield conversion
# 2 by 2 design, replicated 3 times

# create the data
Reactant <- rep(c("low","high"), times = 6)
Catalyst <- rep(c("low","low","high","high"), times = 3)
Replication <- rep(c("1","2","3"), each = 4)
YieldConvertion <- c(28,36,18,31,25,32,19,30,27,32,23,29)
ChemicalData <- data.frame(Reactant, Catalyst, Replication, YieldConvertion)
```

```{r}
# carry out ANOVA using CRD
ChemModel <- aov(YieldConvertion ~ Reactant * Catalyst, data = ChemicalData)
# model summary
summary(ChemModel)
```

* The main effect of Reactant is statistically significant and large (F(1, 8) = 53.19, p < .001)
* The main effect of Catalyst is also statistically significant and large (F(1, 8) = 19.15, p = 0.002)
* The interaction between Reactant and Catalyst is not statistically significant (F(1, 8) = 2.13, p = 0.183)

```{r}
# carry out ANOVA using RBD
ChemModel_2 <- aov(YieldConvertion ~ Replication + Reactant * Catalyst, data = ChemicalData)
# model summary
summary(ChemModel_2)
```

* I used Replications as blocks and they don't differ significantly at 5% level of significance (Of course replications aren't suppose to differ). Replications are used to attain precision.
* The main effect of Reactant is statistically significant and large (F(1,6) = 50.34, p < .001).
* The main effect of Catalyst is also statistically significant and large (F(1,6) = 18.12, p = 0.005).
* The interaction between Reactant and Catalyst is not statistically significant (F(1, 6) = 2.01, p = 0.206).

```{r}
# post-hoc multiple range test using Tukey's HSD test
TukeyHSD(ChemModel_2)
```

* The Tukey's HSD test confirms that the replications are not statistically different and this is good.
* There are significant differences between the levels of reactant and catalyst.
* All the interaction effects are statistically significant, except (high Reactant amount at low Catalyst amount vs high Reactant amount at high Catalyst amount) and (low Reactant amount at low Catalyst amount vs high Reactant amount at high Catalyst amount).

```{r}
## Assess if there is a statistical difference in the quantity of Dirt removed when two detergents are used, each at different temperatures

# Another 2 by 2 factorial design

# first create the data
Temperature <- rep(c("warm","hot"), each = 5, times = 2)
Detergent <- rep(c("X","Y"), each = 10)
Dirt_removed <- c(14,16,15,12,17,18,19,17,19,20,
                  17,19,20,17,22,21,20,18,22,23)
DirtRemovalData <- data.frame(Detergent, Temperature, Dirt_removed)
```
```{r}
# carry out ANOVA
DirtModel <- aov(Dirt_removed ~ Detergent * Temperature, data = DirtRemovalData)
# model summary
summary(DirtModel)
```

* The main effect of Detergent is statistically significant and large (F(1,16) = 15.52, p = 0.001).
* The main effect of Temperature is statistically significant and large (F(1, 16) = 11.88, p = 0.003).
* The interaction between Detergent and Temperature is not statistically significant (F(1,16) = 1.52, p = 0.236).

```{r}
# post-hoc multiple range test using Tukey's HSD test
TukeyHSD(DirtModel)
```

* There are significant differences between the two detergents and the three temperature levels with regards to the quantity of dirt removed.
* Detergent X when used at warm temperature and also at hot temperature removes significantly different quantity of dust.
* Detergent X when used at warm temperature vs detergent Y when used at hot temperature, removes significantly different quantity of dust.
* Detergent Y when used at warm temperature vs Detergent X when used at warm temperature, removes significantly different quantity of dust.

```{r}
## Ultra Filtration Experiment (2 by 2 replicated 4 times)

# first create a data frame with the factors and values
Additive <- rep(c("low","high"),each = 4, times = 2)
Time <- rep(c("1","3"), each = 8) # the no. of times additive was applied
Quantity <- c(69.6,71.5,70,69,91,93,93.2,87.2,80,81.6,83,84.3,92.3,93.4,
                 88.5,95.6)
UltraData <- data.frame(Additive, Time, Quantity)
```
```{r}
# carry out ANOVA
UltraFiltrationModel <- aov(Quantity ~ Additive * Time, data = UltraData)
summary(UltraFiltrationModel)
```

* The main effect of Additive is statistically significant and large (F(1,12) = 185.59, p < .001).
* The main effect of Time is statistically significant and large (F(1, 12) = 34.78, p < .001).
* The interaction between Additive and Time is statistically significant and large (F(1,12) = 22.30, p < .001).

```{r}
# post-hoc multiple range test using Tukey's HSD test
TukeyHSD(UltraFiltrationModel)
```

* There are significant differences between the amount of additive applied and the number of times the additive was applied with regards to the quantity of ultra-filtration achieved.
* High amount of additive when applied 3 times and also when applied only once, doesn't result in a significant difference in quantity of ultra-filtration achieved.
* All the remaining interaction effects differ significantly with regards to the quantity of ultra-filtration achieved.


# 5. Split-Plot Design

```{r}
## Assess if pulp preparation method and different temperature levels have significant effect on the yield of paper production

## The experiment was conducted in a split-plot design

# first create the data frame

# create a vector containing block labels
Block <- rep(c("1","2","3"), each = 12)
# create a vector containing the different types of pulp
Pulp <- rep(c("1","2","3"), times = 12)
# create a vector containing the different temperature levels
Tmp <- rep(c("200","225","250","275"), each = 3, times = 3)
# create a vector containing the paper yields
paper_yield <- c(30,34,29,35,41,26,37,38,33,36,42,36,28,31,31,32,36,30,
                 40,42,32,41,40,40,31,35,32,37,40,34,41,39,39,40,44,45)

# combine the vectors into a data frame
PaperData <- data.frame(Block, Pulp, Tmp, paper_yield)
```
```{r}
# carry out ANOVA
PaperModel <- sp.plot(block = Block, pplot = Pulp, splot = Tmp, Y = paper_yield)
# have a model summary
summary(PaperModel)
```

* Pulp preparation method has a significant effect on paper yield (F(2, 18) = 9.76, p = 0.0485).
* Temperature levels also has a significant effect on paper yield. (F(3, 18) = 9.76, p < .001).
* The interaction effect between pulp preparation method and the temperature levels is also significant (F(6,18) = 3.15, p = 0.027).

```{r}
# carry out post-hoc multiple comparison tests on treatments using LSD test

# specify the first error degrees of freedom
Error_df_a <- PaperModel$gl.a
Error_df_a
# specify the second error degrees of freedom
Error_df_b <- PaperModel$gl.b
Error_df_b
# specify the first mean square error
MSE_a <- PaperModel$Ea
MSE_a
# specify the second mean square error
MSE_b <- PaperModel$Eb
MSE_b
```
```{r}
# test for difference in means between the three Pulp preparation methods
lsd1 <- LSD.test(y = paper_yield,
                 trt = Pulp,
                 DFerror = Error_df_a,
                 MSerror = MSE_a,
                 alpha = 0.05,
                 p.adj = "bonferroni",
                 group = TRUE,
                 console = TRUE)
lsd1
```

The three types of Pulp preparation methods do not differ significantly with regards to paper yield.

```{r}
# test for difference in mean between the three Temperature levels used
lsd2 <- LSD.test(y = paper_yield,
                 trt = Tmp,
                 DFerror = Error_df_b,
                 MSerror = MSE_b,
                 alpha = 0.05,
                 p.adj = "bonferroni",
                 group = TRUE,
                 console = TRUE)
lsd2
```

* Temperature levels 250 and 275 have almost the same effect on paper yield, but are significantly different from temperature levels 200 and 225.

```{r}
# plot Yield variation by Pulp preparation method
plot(lsd1,
     xlab = "Pulp preparation method",
     ylab = "Yield",
     las = 1
     )
```

There is no difference in yield variations based on the three pulp preparation methods.

```{r}
# plot Yield variation by Temperature level
plot(lsd2,
     xlab = "Temperature level",
     ylab = "Yield",
     las = 1
     )
```

Temperature levels 250 and 275 have almost the same effect on paper yield, but are significantly different from temperature levels 200 and 225.

